\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=0.8in]{geometry}
\usepackage{hyperref}

\parindent 0in

\title{Welcome to Senior Computer Team!}
\author{SCT Officers}
\date{September 25, 2015}

\begin{document}

\maketitle

\section{Introduction}

Welcome to the TJHSST Senior Computer Team. The purpose of this pamphlet is to give you relevant information about all SCT activities.\footnote{The language in this pamphlet is also definitely not ripped from the VMT orientation pamphlet.} We hope to get to know every one of you, and our names are below.
\begin{itemize}
	\item Sponsor: Mr. Rudwick
	\item Captain: Nate Foss
	\item Co-Captains: Ashley Kim, Samuel Hsiang
	\item Statistician: Jason Lam
\end{itemize}
We or any other returning members would be more than happy to answer any questions you may have about SCT, computer science, or school in general.

\section{Eighth Period Structure}

SCT meets every Friday during B block\footnote{Different from A block last year!}. FCT and ICT will meet during A block. Feel free to come as often as you want; however, if you expect to make a travelling team, you should be coming nearly every block.

In previous years we have run two lectures simultaneously during our weekly meetings. We feel that for a computer team we ought to be doing more coding practice. We will be still be split into two levels, one around USACO Gold level, and the other for more introductory material.

However, this year, the two groups will alternate between the lab and the classroom. For example, next week one group will have a full lecture in the classroom, while the other group will have a short lecture preceding coding practice, containing both actual contest problems and problems we write -- anything we feel would be helpful for you. The week after the groups will alternate.

\section{Administrative Business}

If you are not on our mailing list, please let us know ASAP.  We send out announcements and lectures via this list, so read these. They are short and don't bite, we promise. We have had issues of our emails being sent to the spam in the past. Make sure this is not happening for you.

If you do not have a Codeforces\footnote{\url{http://codeforces.com/}} (CF) account, or are not in the TJHSST CF group, make one and join.

If you are not in the TJHSST SCT Facebook group, please join that, too. We post announcements there frequently.

If you have any questions or concerns, email us at \url{tjhsstsctcaptains@gmail.com}.

\begin{center}
\Large{\url{https://activities.tjhsst.edu/sct/}}
\end{center}

\eject

\section{Learn}

For the first day, we've found some practice for you to do on Codeforces. You can find it here:

\begin{center}
\Large{\url{http://codeforces.com/group/M4wsRWBHyZ/contest/204396}}
\end{center}

Also (shameless plug) Samuel put together some stuff you should totally look at:

\begin{center}
\Large{\url{https://www.dropbox.com/s/z2lur71042pjaet/guide.pdf?dl=0}}
\end{center}

Other very useful resources can be found on our website.

Looking forward to the coming year!

\end{document}